---
title: "Comparison between different colormaps for element mapping"
author: "Max Hohl"
date: "22/10/2021"
output:
  bookdown::html_document2:
    toc: yes
    fontsize: 16
    toc_float: yes
    df_print: paged
    fig.cation: yes
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

# Introduction

This notebook highlights the problems that colorblind people are facing by using common colormaps, such as "Rainbow" or "Heat", for element maps of minerals. Plotting mineral maps "through the eyes of colorblind people" shows that the same color occurs at least twice in a given range, representing completely different values. As shown below, those issues, which puts roughly xx% of the population at a disadvantage, can be mitigated by using colormaps like `viridis` that designed in way that those color repetition don't occur. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

I tend to only load the tidyverse packages at the beginning of a notebook.
Functions from other packages are loaded explicitly. This helps me to remember
which function is from which package.

```{r}
library(tidyverse)
# require(cowplot)
# require(colorblindcheck)
```

Install required packages for the this project. `colorspace` and `colorblindr`
simulate the colormaps the way colorblind people see them. The `geochem` package
includes plotting functions for element maps.

```{r}
# install.packages("colorspace", repos = "http://R-Forge.R-project.org")
# remotes::install_github("clauswilke/colorblindr")
# remotes::install_github("muhohl/geochem")
```

# Comparison of Colormaps

A simple comparison of the three different colormaps and how they are perceived
by colorblind people.

## Rainbow 

```{r}
rainbow_palette <- rev(c("#CCCCCC", "#8B0000", "#FF0000", "#FFA500",
                     "#FFFF00", "#00FF00", "#90EE90",
                     "#ADD8E6", "#00008B",  "#A020F0", "#000000"))
colorblindcheck::palette_check(rainbow_palette, plot = TRUE, )
```

The popular Rainbow colormap. Both for Protanopia and Tritanopia black colors
occur at least twice in each color range.

## Heat

```{r}
heat_palette <- rev(c("#FFFF00", "#FFA500", "#FF0000", "#8B0000", "#000000"))
colorblindcheck::palette_check(heat_palette, plot = TRUE)
```

Another common colormap using yellow for high values, for low values red and
the lowest are in black. Red and black colors are difficult to distinguish for
people suffering from Deuteranopia and Protanopia.


## Viridis (C)

```{r}
colorblindcheck::palette_check(viridis::viridis(11), plot = TRUE)
```

Example of the viridis color scale shows clearly distinguishable colors for all
three types of colorblindness. The viridis colormap includes four different
scales (A, B, C, D), with the here shown C type being the most popular one.

# Element Maps

Loading the data frame, containing the image data.

```{r}
df_element_map <- read_csv("Image_data.csv")
```

Comparison between different colormaps for selected elements. 

## Rainbow Map

```{r}
i <- 1
rainbow_element_maps <- list()
for (element in names(df_element_map[-c(1:2)])) {
    p.rainbow <- ggplot(data = df_element_map,
                aes(x, y,
                    fill = !!sym(element))) +
    geom_raster(interpolate = TRUE) +
    coord_fixed(ratio = 1) +
    scale_y_discrete(expand = c(0,0)) +
    scale_x_discrete(expand = c(0,0)) +
    labs(fill = "",
         y = "",
         x = "") +
    ggtitle(paste(element, "ppm")) +
    scale_fill_gradientn(colors = rainbow_palette,
                         trans = "log",
                         breaks = c(1, 10, 100, 1000, 10000),
                         labels = scales::label_scientific()) +
    theme_bw() +
    theme(panel.border = element_blank(),
          text = element_text(family = "serif",
                              size = 16)) +
    guides(fill = guide_colorbar(barwidth = unit(1, "lines"),
                                 barheight = unit(8, "lines"),
                                 ticks.colour = "black", 
                                 frame.colour = "black"))
    rainbow_element_maps[[i]] <- p.rainbow
    i <- i+1
}
p.g.rainbow <- cowplot::plot_grid(plotlist = rainbow_element_maps)

```

```{r, fig.width=12, fig.height=7}
p.g.rainbow
```

## Heat Map

```{r}
i <- 1
heat_element_maps <- list()
for (element in names(df_element_map[-c(1:2)])) {
    p.rainbow <- ggplot(data = df_element_map,
                aes(x, y,
                    fill = !!sym(element))) +
    geom_raster(interpolate = TRUE) +
    coord_fixed(ratio = 1) +
    scale_y_discrete(expand = c(0,0)) +
    scale_x_discrete(expand = c(0,0)) +
    labs(fill = "",
         y = "",
         x = "") +
    ggtitle(paste(element, "ppm")) +
    scale_fill_gradientn(colors=heat_palette,
                         trans = "log",
                         breaks = c(1, 10, 100, 1000, 10000),
                         labels = scales::label_scientific()) +
    theme_bw() +
    theme(panel.border = element_blank(),
          text = element_text(family = "serif",
                              size = 16)) +
    guides(fill = guide_colorbar(barwidth = unit(1, "lines"),
                                 barheight = unit(8, "lines"),
                                 ticks.colour = "black", 
                                 frame.colour = "black"))
    heat_element_maps[[i]] <- p.rainbow
    i <- i+1
}
p.g.heat <- cowplot::plot_grid(plotlist = heat_element_maps)
```

```{r, fig.width=12, fig.height=7}
p.g.heat
```

## Viridis Map (C)

```{r, fig.width=12, fig.height=7}
viridis_d_element_maps <- 
    geochem::laser_map(data = df_element_map, 
                       selected_elements = names(df_element_map[-c(1:2)]),
                       fontsize = 16)
cowplot::plot_grid(plotlist = viridis_d_element_maps)
```

## Viridis Map (B)

```{r, fig.width=12, fig.height=7}
viridis_b_element_maps <- 
    geochem::laser_map(data = df_element_map, 
                       selected_elements = names(df_element_map[-c(1:2)]),
                       option = "B", fontsize = 16)
cowplot::plot_grid(plotlist = viridis_b_element_maps)
```

## Colorblind Rainbow 

```{r, fig.height=8, fig.width=10}
cowplot::plot_grid(rainbow_element_maps[[4]],
                   colorblindr::cvd_grid(plot = rainbow_element_maps[[4]]),
                   nrow = 2, rel_heights = c(0.33, 0.67))
```



```{r, fig.height=8, fig.width=10}
cowplot::plot_grid(rainbow_element_maps[[5]],
                   colorblindr::cvd_grid(plot = rainbow_element_maps[[5]]),
                   nrow = 2, rel_heights = c(0.33, 0.67))
```

Concentration of As are very low in the core of the pyrite and high in the
zonation of the rim.

## Colorblind Heat

```{r, fig.height=8, fig.width=10}
cowplot::plot_grid(heat_element_maps[[5]],
                   colorblindr::cvd_grid(plot = heat_element_maps[[5]]),
                   nrow = 2, rel_heights = c(0.33, 0.67))

```

```{r, fig.height=8, fig.width=10}
cowplot::plot_grid(viridis_b_element_maps[[5]],
                   colorblindr::cvd_grid(plot = viridis_b_element_maps[[5]]),
                   nrow = 2, rel_heights = c(0.33, 0.67)
)
```

## Colorblind Viridis (C)
```{r, fig.height=8, fig.width=10}
cowplot::plot_grid(viridis_d_element_maps[[4]],
                   colorblindr::cvd_grid(plot = viridis_d_element_maps[[4]]),
                   nrow = 2, rel_heights = c(0.33, 0.66)
                   )
```




```{r, fig.height=8, fig.width=10}
cowplot::plot_grid(viridis_d_element_maps[[5]],
                   colorblindr::cvd_grid(plot = viridis_d_element_maps[[5]]),
                   nrow = 2, rel_heights = c(0.33, 0.66)
                   )
```

# Session Info

```{r}
sessionInfo()
```

